Learning, Evaluation and Recognition of \strips\ Action Models with Classical Planning

### 2018-04-16 doc/aij18/aij18.tex
updates added to section4 AIJ18 paper

Last but not least, we say that two \strips\ operator schemes $\xi$ and $\xi'$ are {\em comparable} if both schemas have the same headers so they can be built from the same set of possible elements. Formally, iff $head(\xi)=head(\xi')$ and it holds that $F_v(\xi)=F_v(\xi')$. For instance we can claim that the {\tt stack} and {\tt unstack} blocksworld operators are {\em comparable} while  {\tt stack} and {\tt pickup} are not. Likewise we say that two \strips\ action models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there is a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every $\xi\in\mathcal{M}$ to a comparable action schema $\xi'\in\mathcal{M'}$ and viceversa.

\section{Learning \strips\ action models}
\label{sec:Section4}
Learning \strips\ action models from fully available input knowledge, i.e. from plans where the {\em pre-} and {\em post-states} of every action are available, is straightforward~\cite{jimenez2012review}:
\begin{itemize}
  \item {\em Preconditions} are derived lifting the minimal set of literals that appears in all the pre-states of the corresponding actions. 
  \item {\em Effects} are derived lifting the literals that change between the pre and post-state of the corresponding action executions.
\end{itemize}
This section formalizes more challenging learning tasks, where less input knowledge is available.

\subsection{Learning from observations of plan executions}
The first learning task corresponds to observing an agent acting in the world but watching only the states that result of its actions, the actual executed actions are unobserved. This learning task is formalized as a tuple $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$:
\begin{itemize}
\item $\mathcal{M}$, the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. 
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is the sequence of {\em state observations} obtained watching the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. We assume that the initial state $s_0$ is {\em fully observable} while the rest of states in $\mathcal{O}$ can be {\em partially observable}, meaning that some of the observed fluents are missing because it is unknown whether their value is either positive or negative. In the extreme, full state observations could be missing but we assume that state observations are {\em noiseless}, meaning that if the value of a fluent is observed it is correct. 
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. Note that $\Psi$ can be inferred from the state observations provided that at least a state $s\in \mathcal{O}$ is a full state, that is $|s|=|F|$.
\end{itemize}

A solution to the $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ learning task is a set of operator schema $\mathcal{M}'$ that is compliant with the input model $\mathcal{M}$, the given state observations $\mathcal{O}$ and the predicates $\Psi$. Solving this learning task implies determining, not only the \strips\ action model $\mathcal{M}'$, but also the unobserved plan $\pi$, that explains the input observations with the learned \strips\ model. Figure~\ref{fig:example-observations} shows an example of a $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ task for learning the {\em blocksworld} \strips\ action model from the five-state observations sequence that corresponds to inverting a 2-block tower. Learning the action model from this example implies inferring that the unobserved plan is $\pi=\tup{\small\tt (unstack\ B\ A), (putdown\ B), (pickup\ A), (stack\ A\ B)}$.

@@

In some cases we may not require to start learning from scratch because some preconditions and/or some positive or negative effects may be a priori known. The operator schemas in $\mathcal{M}$, that are given as input, may be not {\em empty} but {\em partially specified}. Such scenario is relevant for {\em Policy learning}. A policy is function that maps states into actions and represents the conditions under which actions should be applied to achieve certain goals. Given an action model $\mathcal{M}$ the task of learning a policy that is compliant with a given set of observations of plan executions can be defined as the task of learning extra preconditions for each action scheme $\xi\in\mathcal{M}$. These extra preconditions capture when actions can be applied according to the policy. A policy that defines a single applicable action for each reachable state is a {\em full policy} otherwise, we say it is a {\em partial policy}. In the general case, learning a {\em full policy} for an arbitrary planning task is complex because it may require computing {\em high-level state features}~\cite{lotinac2016automatic}.

Here we redefine the learning task to cover the scenario where some of the actions executed by the observed agent are available. If all executed actions are known, $\mathcal{O}$ should contain partial states because otherwise, the learning task becomes trivial as explained. The learning task is now formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is the action sequence that produces the sequence of state observations given in $\mathcal{O}$. Again we assume that action observations are noiseless, meaning that if the value of an action is observed it is correct. When the input plan is {\em diverse} enough, i.e. $\pi$ contains at least one ground action for each of the aimed action schemes, the set of {\em empty} operator schemas $\mathcal{M}$ can be inferred from $\pi$.
\end{itemize}

Figure~\ref{fig:example-plans} shows an example of a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, that corresponds to observing the execution of a four-action plan for inverting a two-block tower. The exaple skips $\mathcal{M},\Psi$ since they are the same as in Figure~\ref{fig:example-observations}. In this example $\mathcal{O}=\tup{s_0,s_4}$ which means that only the first and last states were observed and the three intermediate states $s_1$, $s_2$ and $s_3$ are fully unknown.

@@

The previous definitions formalize the learning of \strips\ action models from the observation of a single plan execution. These definitions are extensible to the more general case where the available input data is obtained from the execution of multiple plans:
\begin{itemize}
\item When learning from states observations, $\Lambda=\tup{\mathcal{M},\mathcal{O'},\Psi}$ is still a valid formalization by simply considering that now $\mathcal{O'}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ is a sequence of {\em state observations} obtained watching the execution of a serie of {\em unobserved} plans $\pi^t=\tup{a_1, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$}, one after the other. This means that, for each {\small $1\leq i\leq n^t$}, $a_i^t\in \pi^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
  \item When learning from a set of observed plans, the task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O'},\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is the given sequence of example plans producing the corresponding sequence of state observations $\mathcal{O'}$.
\end{itemize}

>>>>

Last but not least, we say that two \strips\ operator schemes $\xi$ and $\xi'$ are {\em comparable} if both schemas have the same headers so they can be built from the same set of possible elements. Formally, iff $head(\xi)=head(\xi')$ so it also holds that $F_v(\xi)=F_v(\xi')$. For instance we can claim that the {\tt stack} and {\tt unstack} blocksworld operators are {\em comparable} while  {\tt stack} and {\tt pickup} are not. Likewise we say that two \strips\ action models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every $\xi\in\mathcal{M}$ to a comparable action schema $\xi'\in\mathcal{M'}$ and viceversa.

\section{Learning \strips\ action models}
\label{sec:Section4}
Learning \strips\ action models from fully available input knowledge, i.e. from plans where the {\em pre-} and {\em post-states} of every action are available, is straightforward~\cite{jimenez2012review}:
\begin{itemize}
  \item {\em Preconditions} are derived lifting the minimal set of literals that appears in all the pre-states of the corresponding action, that is any action that belongs to the same operator scheme. 
  \item {\em Effects} are derived lifting the literals that change between the pre and post-state of the corresponding action executions.
\end{itemize}
This section formalizes more challenging learning tasks where less input knowledge is available for instance, because it cannot be observed.

\subsection{Learning from observations of plan executions}
The first learning task corresponds to observing an agent acting in the world but watching only the states that result of its actions, the actual executed actions are unobserved. This learning task is formalized as a tuple $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$:
\begin{itemize}
\item $\mathcal{M}$, the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. 
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is the sequence of {\em state observations} obtained watching the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. We assume that the initial state $s_0$ is {\em fully observable} while states $s_i$ s.t. {\small $1\leq i\leq |\mathcal{O}|$} can be {\em partially observable}, meaning that some fluents in $s_i$ are missing because it is unknown whether their value is either positive or negative. In the extreme, states $s_i$ {\small $1\leq i\leq |\mathcal{O}|$}, can be missing but we assume that any state observation is {\em noiseless}, meaning that if the value of a fluent is observed it is correct. 
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. Note that $\Psi$ can be inferred from the state observations provided that at least a state $s\in \mathcal{O}$ is a full state, that is $|s|=|F|$.
\end{itemize}

A solution to a $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ learning task is a set of operator schema $\mathcal{M}'$ that is compliant with the input model $\mathcal{M}$, the given state observations $\mathcal{O}$ and the predicates $\Psi$. Solving $\Lambda$ implies determining, not only the \strips\ action model $\mathcal{M}'$, but also the unobserved plan $\pi$, that explains the input observations with the learned \strips\ model. Figure~\ref{fig:example-observations} shows an example of a $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ task for learning the {\em blocksworld} \strips\ action model from the five-state observations sequence obtained inverting a 2-block tower. Learning the action model from this example implies inferring the unobserved plan $\pi=\tup{\small\tt (unstack\ B\ A), (putdown\ B), (pickup\ A), (stack\ A\ B)}$.

@@

In some cases we may not require to start learning from scratch because some preconditions and/or some positive or negative effects may be a priori known. The operator schemas in $\mathcal{M}$, that are given as input, may be not {\em empty} but {\em partially specified}. Such scenario is relevant for {\em policy learning}. A policy is function that maps states into actions and represents the conditions under which actions should be applied to achieve certain goals. Given an action model, the task of learning a policy that is compliant with a set of observations can be defined as learning extra preconditions for each action scheme in the model. These extra preconditions capture when actions can be applied according to the policy. A policy that exactly defines a single applicable action for each reachable state is a {\em full policy} otherwise, we say it is a {\em partial policy}. In the general case, learning a {\em full policy} for an arbitrary planning task is complex because a given action scheme may be applicable in different situations and also because the compact representation of this set of different situations may require the computation of {\em high-level state features}~\cite{lotinac2016automatic}.

The $\Lambda$ learning task can also be redefined to cover the scenario where some of the actions executed by the observed agent are available. If all the executed actions are known then states observations should be partial otherwise, the learning task is trivial. The learning task is now formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is the action sequence that produces the sequence of state observations given in $\mathcal{O}$. Again we assume that action observations are noiseless, meaning that if the value of an action is observed it is correct. When the input plan is {\em diverse} enough, i.e. $\pi$ contains at least one ground action for each of the aimed action schemes, the set of {\em empty} operator schemas $\mathcal{M}$ can be inferred from $\pi$.
\end{itemize}

Figure~\ref{fig:example-plans} shows an example of a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, that corresponds to observing the execution of a four-action plan for inverting a two-block tower. In this example $\mathcal{O}=\tup{s_0,s_4}$ which means that only the first and last states are observed and the three intermediate states $s_1$, $s_2$ and $s_3$ are fully unknown. $\mathcal{M},\Psi$ are skipped, since they are the same as in Figure~\ref{fig:example-observations}.

@@

The previous definitions formalize the learning of \strips\ action models from the observation of a single plan execution. These definitions are extensible to the more general case where the execution of multiple plans is observed. The task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O'},\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is the given sequence of example plans producing the corresponding sequence of state observations $\mathcal{O'}$. Now $\mathcal{O'}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ is a sequence of {\em state observations} obtained watching the execution of a serie of {\em unobserved} plans $\pi^t=\tup{a_1, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$}, one after the other. This means that, for each {\small $1\leq i\leq n^t$}, $a_i^t\in \pi^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
